---
title: "Determining The Odds For Cardiovascular Disaease"
subtitle: "Using Framingham Heart Study Data"
author: "By Matthew Curcio"
geometry: margin=1.9cm
output: pdf_document
---

    
## 1. Executive Summary

This report investigates data from the 1948 Framingham Heart Study.^[https://www.framinghamheartstudy.org] Data can be found at Kaggle.^[https://www.kaggle.com/datasets/captainozlem/framingham-chd-preprocessed-data]

Using this data, I investigate the risk factors and odds for seven factors that increase the likelihood of cardiovascular disease (CVD).


1. This report and my article Introduction to Logit^[https://github.com/mcc-us/intro-2-logit/blob/main/intro-2-logit.pdf] discuss ideas concerning logistic regression and R. 


2. Seven (7) factors have a strong correlation that leads to cardiovascular disease. The odds related to each factor are calculated from the data.

### Seven Factors and YOUR Odds Of Developing Cardiovascular Disease

| No. | Factors | Approximate Odds |
| :-- | :------ | ----------------:|
| 1   | Age (80 yr Male : 20 yr Male)          | 28.0:1  |
| 2   | High Systolic Blood Pressure           | 7.8:1 |
| 3   | High Glucose Levels                    | 2.5:1 |
| 4   | Prevalence Of Stroke In Family History | 2.4:1 |
| 5   | Cigarettes Per Day                     | 2.1:1 |
| 6   | Male Vs Female                         | 1.5:1 |
| 7   | Prevalence Of Hypertension In Family History | 1.3:1 |


## 2. Introduction
    
Odds are calculated by using dividing two probabilities. In the case of a *Risk of Stroke,* divide the Probability *WITH* the prevalence of stroke in a patients' family history *VERSUS* the Probability *WITHOUT* having the prevalence of stroke in the family history.

*For Example:*

$$Odds~of~Stroke ~~=~~ \frac{Probability~[Stroke~in~family~history]}{Probability~[NO~stroke~in~family~history]} ~=~ \frac{0.18761}{0.07778} ~=~ \frac{2.41}{1}$$

This longitudinal study includes 4,133 participants with 13 factors total over 10 years. 

>The Framingham Heart Study began in 1948 by recruiting 5,209 men and women between the ages of 30 and 62 from the town of Framingham, Massachusetts. [These recruits] had not yet developed overt symptoms of cardiovascular disease or suffered a heart attack or stroke. 
>
>[The] study has since led to the identification of major CVD risk factors, as well as valuable information on the effects of these factors such as blood pressure, blood triglyceride and cholesterol levels, age, gender, and psychosocial issues.^[https://www.framinghamheartstudy.org/fhs-about]

\newpage

## 3. Exploratory Data Analysis

```{r LoadData, message=FALSE, warning=FALSE, include=FALSE}
# DO NOT SHOW IN FINAL REPORT
options(tinytex.verbose = FALSE)
options(digits = 5) 

#Load Libraries
Libraries <- c("knitr", "readr", "ggplot2")
for (p in Libraries) { 
    library(p, character.only = TRUE)
}

#Load Data
df <- read_csv("data/CHD_preprocessed.csv", 
     col_types = cols(male = col_factor(levels = c("0", "1")), 
                      education = col_factor(levels = c("0", "1")),
                      currentSmoker = col_skip(), 
                      BPMeds = col_skip(), 
                      prevalentStroke = col_factor(levels = c("0", "1")), 
                      prevalentHyp = col_factor(levels = c("0", "1")), 
                      diabetes = col_factor(levels = c("0", "1")), 
                      TenYearCHD = col_factor(levels = c("0", "1"))))
```


### Descriptive Statistics

#### Age Of Participants, Cigarettes Per Day, Body Mass Index

```{r}
par(mfrow = c(1, 3))

boxplot(df$age, ylim=c(0,70), col = "red")
title('Age Of Participants')

boxplot(df$cigsPerDay, ylim=c(0,70), col = "#EEEEEC")
title('Cigarettes Per Day')

boxplot(df$BMI, ylim=c(0,70), col = "blue")
title('Body Mass Index')
```



## 2. Logistic Regression Model Results

### 2.1 Model Using 13 Factors

```{r cache=TRUE}
mylogit <- glm(TenYearCHD ~ male + age + education + cigsPerDay + glucose +
                            prevalentStroke + prevalentHyp + diabetes + 
                            totChol + sysBP + diaBP + BMI + heartRate,
               data = df, 
               family = "binomial")

summary(mylogit)
```

- **7 most significant variables**

- Seven predictors have $\small \alpha < 0.05$. They are significant and associated with acquiring cardiovascular disease.


| Rank | Risk Factor                 |
| :--- | :-------------------------- |
| 1    | Prevalence of Stroke1       |
| 2    | Male1                       |
| 3    | Prevalence of Hypertension1 |
| 4    | Age                         |
| 5    | Cigarettes Per Day          |
| 6    | Systolic Blood Pressure     |
| 7    | Glucose                     |


<br>

### 2.2 Wald Test: Do The Seven Factors Fit Our Model

- The Wald Chi-Square Test can help determine if our proposed model is significant.

- The Wald test generates a *P-value << 0.001*. 

- Therefore, we conclude the seven (7) parameters are significant and useful in describing cardiovascular disease.

```{r eval=FALSE, cache=TRUE, include=FALSE}
library(aod)

wald.test(b = coef(mylogit), 
          Sigma = vcov(mylogit), 
          Terms = c(2, 3, 5, 6, 7, 10, 14))
```

<br>

### 2.3 Determination of Odds for Seven Variables

- We can calculate the odds of acquiring cardiovascular disease for each of the seven variables. 

- By holding all other values constant we create a dataframe that investigates the odds given Prevalence of Stroke, for example.

```{r Stroke, cache=TRUE}
strok_test <- with(df, data.frame(male = "0",
                                  age = mean(age),
                                  education = "0",
                                  cigsPerDay = 0, # Non-smoker
                                  prevalentHyp = "0",
                                  diabetes = "0",
                                  totChol = mean(totChol),
                                  sysBP = mean(sysBP),
                                  diaBP = mean(diaBP),
                                  BMI = mean(BMI),
                                  heartRate = mean(heartRate),
                                  glucose = mean(glucose),
                                  prevalentStroke = c("0", "1")) 
                   )

# Convert prevalentStroke from Numeric to FACTOR
strok_test$prevalentStroke <- as.factor(strok_test$prevalentStroke) 
# str(strok_test)

strok_test$prevalentStroke <- predict(mylogit, 
                                      newdata = strok_test, 
                                      type = "response")
#strok_test$prevalentStroke
```


### 2.4 **Odds Given Prevalence Of Stroke In family history.**

1. WITH Prevalence of Stroke: `r strok_test$prevalentStroke[[2]]`
2. NO Prevalence of Stroke: `r strok_test$prevalentStroke[[1]]`

- **Odds = `r strok_test$prevalentStroke[[2]]/strok_test$prevalentStroke[[1]]`**

<br>

### 2.5 **Odds Given For Male Vs Female**

```{r MF, cache=TRUE}
male_test <- with(df, data.frame(male = c("0","1"), # Factor of Interest
                                 age = mean(age),
                                 education = "0",
                                 cigsPerDay = 0,
                                 prevalentHyp = "0",
                                 diabetes = "0",
                                 totChol = mean(totChol),
                                 sysBP = mean(sysBP),
                                 diaBP = mean(diaBP),
                                 BMI = mean(BMI),
                                 heartRate = mean(heartRate),
                                 glucose = mean(glucose),
                                 prevalentStroke = "0"))

# REMEMBER convert male_test from numeric to FACTOR
male_test$male <- as.factor(male_test$male) 

male_test$male <- predict(mylogit, newdata = male_test, type = "response")
```

1. Males: `r male_test$male[[2]]`
2. Female: `r male_test$male[[1]]` 

- **Odds = `r male_test$male[[2]]/male_test$male[[1]]`**

<br>

### 2.6 Odds Prevalence of Hypertension In Family History


```{r hyperT, cache=TRUE}
hyperT_test <- with(df, data.frame(male = "0",
                                   age = mean(age),
                                   education = "0",
                                   cigsPerDay = 0,
                                   prevalentHyp = c("0","1"), # Factor of Interest
                                   diabetes = "0",
                                   totChol = mean(totChol),
                                   sysBP = mean(sysBP),
                                   diaBP = mean(diaBP),
                                   BMI = mean(BMI),
                                   heartRate = mean(heartRate),
                                   glucose = mean(glucose),
                                   prevalentStroke = "0"))

# REMEMBER convert male_test from numeric to FACTOR
hyperT_test$prevalentHyp <- as.factor(hyperT_test$prevalentHyp) 

hyperT_test$prevalentHyp <- predict(mylogit, newdata = hyperT_test, type = "response")
```


1. WITH Prevalence of Hypertension: `r hyperT_test$prevalentHyp[[2]]` 
2. NO Prevalence of Hypertension: `r hyperT_test$prevalentHyp[[1]]` 

- **Odds = `r hyperT_test$prevalentHyp[[2]]/hyperT_test$prevalentHyp[[1]]`**

<br>

### 2.7 Odds Given Age


```{r Age, cache=TRUE}
age_test <- with(df, data.frame(male = "0",
                                   age = c(20,30,40,50,60,70,80),
                                   education = "0",
                                   cigsPerDay = 0,
                                   prevalentHyp = "0",
                                   diabetes = "0",
                                   totChol = mean(totChol),
                                   sysBP = mean(sysBP),
                                   diaBP = mean(diaBP),
                                   BMI = mean(BMI),
                                   heartRate = mean(heartRate),
                                   glucose = mean(glucose),
                                   prevalentStroke = "0"))

age_test$age <- predict(mylogit, newdata = age_test, type = "response")
```


| Age (years) | Probability Given Age |       Odds Compared to 20 yr old        |
| ----------: | :-------------------: | :-------------------------------------: |
|          20 | `r age_test$age[[1]]` | `r age_test$age[[1]]/age_test$age[[1]]` |
|          30 | `r age_test$age[[2]]` | `r age_test$age[[2]]/age_test$age[[1]]` |
|          40 | `r age_test$age[[3]]` | `r age_test$age[[3]]/age_test$age[[1]]` |
|          50 | `r age_test$age[[4]]` | `r age_test$age[[4]]/age_test$age[[1]]` |
|          60 | `r age_test$age[[5]]` | `r age_test$age[[5]]/age_test$age[[1]]` |
|          70 | `r age_test$age[[6]]` | `r age_test$age[[6]]/age_test$age[[1]]` |
|          80 | `r age_test$age[[7]]` | `r age_test$age[[7]]/age_test$age[[1]]` |

<br>


### 2.8 Odds Given Number Of Cigarettes Per Day

```{r Cigarettes, cache=TRUE}
cigs_per_day <- with(df, data.frame(male = "0",
                                    age = mean(age),
                                    education = "0",
                                    cigsPerDay = c(0,10,20,30,40),
                                    prevalentHyp = "0",
                                    diabetes = "0",
                                    totChol = mean(totChol),
                                    sysBP = mean(sysBP),
                                    diaBP = mean(diaBP),
                                    BMI = mean(BMI),
                                    heartRate = mean(heartRate),
                                    glucose = mean(glucose),
                                    prevalentStroke = "0"))

cigs_per_day$cigsPerDay <- predict(mylogit, newdata = cigs_per_day, type = "response")

#cigs_per_day$cigsPerDay
```

1. A pack of cigarettes gave a person 45% increase of acquiring Cardiovascular disease, **using this data set. This seems oddly low.**

| Age (years) |      Probability Given Age       |                  Odds Compared to Zero Cigarettes Per Day     |
| ----------: | :------------------------------: | :-----------------------------------------------------------: |
|           0 | `r cigs_per_day$cigsPerDay[[1]]` | `r cigs_per_day$cigsPerDay[[1]]/cigs_per_day$cigsPerDay[[1]]` |
|          10 | `r cigs_per_day$cigsPerDay[[2]]` | `r cigs_per_day$cigsPerDay[[2]]/cigs_per_day$cigsPerDay[[1]]` |
|          20 | `r cigs_per_day$cigsPerDay[[3]]` | `r cigs_per_day$cigsPerDay[[3]]/cigs_per_day$cigsPerDay[[1]]` |
|          30 | `r cigs_per_day$cigsPerDay[[4]]` | `r cigs_per_day$cigsPerDay[[4]]/cigs_per_day$cigsPerDay[[1]]` |
|          40 | `r cigs_per_day$cigsPerDay[[5]]` | `r cigs_per_day$cigsPerDay[[5]]/cigs_per_day$cigsPerDay[[1]]` |

<br>

### 2.9 Odds Given Systolic Blood Pressure

```{r BP, cache=TRUE}
summary(df$sysBP)
# Min. 1st Qu.  Median    Mean  3rd Qu.  Max. 
# 83.5   117.0   128.0   132.4   144.0   295.0 

sysBP_calc <- with(df, data.frame(male = "0",
                                  age = mean(age),
                                  education = "0",
                                  cigsPerDay = 0,
                                  prevalentHyp = "0",
                                  diabetes = "0",
                                  totChol = mean(totChol),
                                  sysBP = c(117, 128, 144, 295),
                                  diaBP = mean(diaBP),
                                  BMI = mean(BMI),
                                  heartRate = mean(heartRate),
                                  glucose = mean(glucose),
                                  prevalentStroke = "0"))

sysBP_calc$sysBP <- predict(mylogit, newdata = sysBP_calc, type = "response")

#sysBP_calc$sysBP
```

   
| Systolic BP | Probability Given Systolic BP |                Odds Systolic BP                 |
| ----------: | :---------------------------: | :---------------------------------------------: |
|         117 |   `r sysBP_calc$sysBP[[1]]`   | `r sysBP_calc$sysBP[[1]]/sysBP_calc$sysBP[[1]]` |
|         128 |   `r sysBP_calc$sysBP[[2]]`   | `r sysBP_calc$sysBP[[2]]/sysBP_calc$sysBP[[1]]` |
|         144 |   `r sysBP_calc$sysBP[[3]]`   | `r sysBP_calc$sysBP[[3]]/sysBP_calc$sysBP[[1]]` |
|    Max  295 |   `r sysBP_calc$sysBP[[4]]`   | `r sysBP_calc$sysBP[[4]]/sysBP_calc$sysBP[[1]]` |
<br>

### 2.10 Odds Given Glucose Levels


```{r Glucose, cache=TRUE}
summary(df$glucose)
#    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
#     40      72      80      82      85     394 

glucose_calc <- with(df, data.frame(male = "0",
                                  age = mean(age),
                                  education = "0",
                                  cigsPerDay = 0,
                                  prevalentHyp = "0",
                                  diabetes = "0",
                                  totChol = mean(totChol),
                                  sysBP = mean(sysBP),
                                  diaBP = mean(diaBP),
                                  BMI = mean(BMI),
                                  heartRate = mean(heartRate),
                                  glucose = c(72, 80, 85, 394),
                                  prevalentStroke = "0"))

glucose_calc$glucose <- predict(mylogit, newdata = glucose_calc, type = "response")

# glucose_calc$glucose.
# 0.094843 0.100852 0.110194 0.239738
```

|  Glucose | Probabilities |  Odds Given Glucose   |
| -------: | :-----------: | :-------------------: |
|       72 |   0.094843    |           1           |
|       80 |   0.100852    | `r 0.100852/0.094843` |
|       85 |   0.110194    | `r 0.110194/0.094843` |
| Max  394 |   0.239738    | `r 0.239738/0.094843` |

<br>


## 3. Conclusion

1. We find seven (7) of the 13 factors lead to cardiovascular disease. The odds related to each factor were calculated from the study.


| No. | Factors | Approximate Odds |
| :-: | :------ | ---------------: |
| 1   | Prevalence Of Stroke In Family History       | 240% |
| 2   | Male Vs Female                               | 150% |
| 3   | Prevalence Of Hypertension In Family History | 130% |
| 4   | Age                                          | 2,800% |
| 5   | Cigarettes Per Day                           | 210% |
| 6   | Systolic Blood Pressure                      | 780% |
| 7   | Glucose Levels                               | 250% |
	

2. The Wald Chi-Square Test can help determine if our proposed model is valuable and significant. The Wald test generates a P-value << 0.001. Therefore, we conclude the seven (7) parameters are significant and useful in describing cardiovascular disease.

3. A pack of cigarettes gave a person 45% increase of acquiring Cardiovascular disease, **using this data set. This seems oddly low.**

| Cigs Per Day | Probability Given Age | Odds Compared to Zero Cigarettes Per Day |
| -: | :----: | :-----------------------------------------------------------: |
| 0 | `r cigs_per_day$cigsPerDay[[1]]` | `r cigs_per_day$cigsPerDay[[1]]/cigs_per_day$cigsPerDay[[1]]` |
| 10 | `r cigs_per_day$cigsPerDay[[2]]` | `r cigs_per_day$cigsPerDay[[2]]/cigs_per_day$cigsPerDay[[1]]` |
| 20 | `r cigs_per_day$cigsPerDay[[3]]` | `r cigs_per_day$cigsPerDay[[3]]/cigs_per_day$cigsPerDay[[1]]` |
| 30 | `r cigs_per_day$cigsPerDay[[4]]` | `r cigs_per_day$cigsPerDay[[4]]/cigs_per_day$cigsPerDay[[1]]` |
| 40 | `r cigs_per_day$cigsPerDay[[5]]` | `r cigs_per_day$cigsPerDay[[5]]/cigs_per_day$cigsPerDay[[1]]` |

---

#### Notes

- For analysis help
https://stats.idre.ucla.edu/r/dae/logit-regression/

- For interpretation help
https://stats.idre.ucla.edu/other/mult-pkg/faq/general/faq-how-do-i-interpret-odds-ratios-in-logistic-regression/.
- https://stats.oarc.ucla.edu/other/mult-pkg/faq/general/faqhow-are-the-likelihood-ratio-wald-and-lagrange-multiplier-score-tests-different-andor-similar/

#### Wald test info
- https://www.mbaskool.com/business-concepts/statistics/6916-wald-test.html
- https://www.statology.org/wald-test-in-r/
- https://handwiki.org/wiki/Wald_test
- https://questionerlab.com/what-is-the-use-of-wald-test-in-logistic-regression
- https://bookdown.org/mike/data_analysis/wald-test.html
- https://bookdown.org/mike/data_analysis/hypothesis-testing.html#wald-test
